---
title       : Top 250 IMDB Movies Visualization
subtitle    : A Shiny application to visualize IMDB Data
author      : Ralph Schlosser
job         : 
logo        : logo.jpg
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

```{r echo=FALSE}
load("..//imdbapp/data/imdb.rda")
```

<style>
.title-slide {
  background-color: #FEFEF5; /* #EDE0CF; ; #CA9F9D*/
}

.title-slide hgroup > h1{
 font-family: 'Oswald', 'Helvetica', sanserif; 
}

.title-slide hgroup > h1, 
.title-slide hgroup > h2 {
  color: #944E2D ;  /* ; #EF5150*/
}

.current article > h2, 
.class hgroup > h2 {
  color: #944E2D ;  /* ; #EF5150*/
}
</style>

## Introduction

1. About the Application
2. Some Challenges
3. Web Scraping with `rvest`

--- .class #id 

## About the Application

* The original idea for the app came from an interesting blog post on [R Bloggers] (http://www.r-bloggers.com/top-250-movies-at-imdb/) about how to retrieve data from the _Internet Movie Data Base_ (IMDB).

* The IMDB contains records of a large number of historic and current movies and television shows.

* It is used and kept up-to-date on a daily basis both by professionals within the movie industry as well as hobbyists. 

* Apart from a variety of information about cast, plot, music, filming locations etc. on a movie, it also maintains a rating index which is derived from IMDB registered users votes.

* This rating is generally seen as measure of quality, i.e. the higher the rating the better the movie.

* The `imdbapp` application visualizes the movie ratings by year, and allows to drill down by rating and country of origin of a particular movie.

--- .class #id 

## Some Challenges

* Since the time the blog mentioned above was written (October 2013), the underlying HTML structure of the IMDB pages have changed.

* This meant that the code used in the blog no longer worked. 

* Most notably, the IMDB has chosen to remove the vote count column.

* The vote count is still displayed, but only on the movie's main page and thus the count had to be retrieved for each movie in the top 250 list individually.

* This was done using a technique that is commonly referred to as "web scraping".

* Also, in order to not upset IMDB due to repeated HTTP request from the same source -- something always to watch out for when scraping the web -- a random delay between 0.5 to 1 second was introduced in the data retrieval function.

--- .class #id 

## Web Scraping with `rvest`

* The `rvest` package (read more on [Rstudio Blogs] (http://blog.rstudio.org/2014/11/24/rvest-easy-web-scraping-with-r/)) was employed for retrieving the required information from the IMDB and build a fresh, top-rated 250 movies data set called `topMovies`.

* When taking a closer look at this data set it appears as if there are far more US movies in the top 250 list, and they've got higher ratings too:

```{r}
table(topMovies$Country[topMovies$Rating > 8.5])
```


